Chemical tagging in lower-dimensional latent space

### 2019-01-22 article/mcfa.tex
lots of stuff -- ready for galah exp4

\shorttitle{A novel approach to chemical tagging in lower-dimensional latent space}
\shortauthors{Casey et al.}

@@

\title{A novel approach to chemical tagging in lower-dimensional latent space}

@@

We performed a grid search for the optimal number of latent factors 
$\NumLatentFactors$ and components $\NumComponents$, starting from 
$\NumLatentFactors = 1$ and $K = 1$, and trialling up until 
$\NumLatentFactors = \todo{8}$ and $\NumComponents = 5$. 
We found that models with higher numbers of latent factors were unable to
converge without adding a regularization constant to the diagonals of the
covariance matrices $\scorecovs$ in latent space. Although it is common
to add a regularization term to the diagonals of covariance matrices
when performing expectation-maximization, we found that our existing
grid search boundaries were sufficient to find the preferred $\NumLatentFactors$
and $\NumComponents$, and chose not to add any regularization term.

@@

Using our preferred model with $J = 5$ and $K = 1$, we sought to find a
rotation matrix that would allow us to identify plausible astrophysical
processes in the factor loads.
Although having to perform this rotation is not comforting, there is utility in
trying to identify the latent factors because the factors can only be rotated by
some valid rotation matrix. That is to say that we will not only exactly what
we have put in: the rotated factors will simply be the closest set of factor
loads that are consistent with the data, allowing us to identify the factor
loads that are astrophysically plausible. We constructed
a $J \times D$ astrophysical factor load matrix $\factorloads_\textrm{A}$ where
all entries were zero except for the following entries:
\begin{itemize}
	\item the entry corresponding to Al in the first latent factor to represent
		  odd-Z element production,
	\item the entries corresponding to Ca and Mg in the second latent factor to
		  represent $\alpha$ particle production,
	\item the entries corresponding to Ni, Co, Fe, Mn, Cr, Ti, and Sc in the third
		  latent factor to represent the Fe-peak group,
	\item the entries corresponding to Sr, Y, and Ba in the fourth latent factor
		  to represent products of the slow neutron capture process, and
	\item the entries corresponding to Eu in the fifth latent factor to represent
		  products of the rapid nucleosynthesis process.
\end{itemize}
These entries were set to $1/\sqrt{E}$ where $E$ is the number of non-zero entries
in that latent factor, in order to ensure that $\factorloads_\textrm{A}$ is
mutually orthogonal. We then found the valid rotation matrix $\vec{R}$ that most
closely approximated the astrophysical loads $\factorloads_\textrm{A}$.
In Figure~\ref{fig:exp2-factor-loads} we show the rotated factor loads found by our
preferred model, where the shaded regions indicate \emph{only} the projected uncertainty on
the rotation angles -- and does not include any uncertainty on the factor loads 
themselves.

@@

The rotated loads are not precisely the same as the target astrophysical loads
in part because the target loads only seek to \emph{identify} and \emph{order}
factors that can be interpreted within an astrophysical context. For example,
the first factor does show positive contributions to Al, like the target load,
but it also shows a slight negative contribution with respect to alpha elements
(Mg, Ca), and a gradual increase among Fe-peak elements up to Ni. Similarly while
Ca and Mg are indeed positive in the second factor, as targeted, there is a
weak negative contribution to Mn abundances, which was not part of the target
loads. The fourth factor shows a positive contributions to Sr, Y, Ba, and to a
lesser extent, Eu, which is representative of slow neutron capture production.
Similarly, the last component shows a very strong positive contribution towards
Eu, and could be interpreted as production from the rapid nucleosynthesis process.

\begin{figure}
	\includegraphics[width=0.5\textwidth]{experiments/exp2-latent-contributions.pdf}
	\caption{Rotated latent factors found in Experiment~2 using the \citet{Barklem:2005}
			 data. \todo{elaborate latent factors}}
    \label{fig:exp2-latent-contribution}
\end{figure}

The relatively wide uncertainty regions on these factors is perhaps not
surprising given that only 61 stars were used in this data set. However, this
experiment hopefully demonstrates that the rotated latent factors have some
interpretability. After accounting for the contributions by all
$J = 5$ factors, the specific scatter remaining in each dimension is shown in
Figure~\ref{fig:exp2-specific-scatter}. The remaining scatter varies from as low 
as 0.01 to 0.02~dex (Mg, Ca, Ti, Mn, Fe, Y) up to 0.08 or 0.16~dex (Al and Ba). 
% TODO: matches Barklem variances?

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp2-latent-factors.png}
	\caption{Rotated latent factors found in Experiment~2 using the \citet{Barklem:2005}
			 data. \todo{elaborate latent factors}}
    \label{fig:exp2-latent-factors}
\end{figure*}

>>>>

\shorttitle{Chemical tagging in lower-dimensional latent space}
\shortauthors{Casey et al.}

@@

\title{Chemical tagging in lower-dimensional latent space}

@@

We performed a grid search for the optimal number of latent factors 
$\NumLatentFactors$ and components $\NumComponents$, starting from 
$\NumLatentFactors = 1$ and $K = 1$, and trialling up until 
$\NumLatentFactors = 7$ and $\NumComponents = 5$. 
We found that models with higher numbers of latent factors were unable to
converge without adding a regularization constant to the diagonals of the
covariance matrices $\scorecovs$ in latent space. Although it is common
to add a regularization term to the diagonals of covariance matrices
when performing expectation-maximization, we found that our existing
grid search boundaries were sufficient to find the preferred $\NumLatentFactors$
and $\NumComponents$, and chose not to add any regularization term.

@@

Using our preferred model with $J = 5$ and $K = 1$, we sought to find a
rotation matrix that would allow us to identify plausible astrophysical
processes in the factor loads.
Although having to perform this rotation is not comforting, there is utility in
trying to identify the latent factors because the factors can only be rotated by
some valid rotation matrix. That is to say that we will not only exactly what
we have put in: the rotated factors will simply be the closest set of factor
loads that are consistent with the data, allowing us to identify the factor
loads that are astrophysically plausible. We constructed
a $J \times D$ astrophysical factor load matrix $\factorloads_\textrm{A}$ where
all entries were zero except for the following entries:
\begin{itemize}
	\item the entry corresponding to Al in the first latent factor to represent
		  odd-Z element production,
	\item the entries corresponding to Ca and Mg in the second latent factor to
		  represent $\alpha$ particle production,
	\item the entries corresponding to Ni, Co, Fe, Mn, Cr, Ti, and Sc in the third
		  latent factor to represent the Fe-peak group,
	\item the entries corresponding to Sr, Y, and Ba in the fourth latent factor
		  to represent products of the slow neutron capture process, and
	\item the entries corresponding to Eu in the fifth latent factor to represent
		  products of the rapid nucleosynthesis process.
\end{itemize}
These entries were set to $1/\sqrt{E}$ where $E$ is the number of non-zero entries
in that latent factor, in order to ensure that $\factorloads_\textrm{A}$ is
mutually orthogonal. We then found the valid rotation matrix $\vec{R}$ that most
closely approximated the astrophysical loads $\factorloads_\textrm{A}$.
In Figure~\ref{fig:exp2-factor-loads} we show the rotated factor loads found by our
preferred model, and illustrate the relative contributions (positive or negative)
to each element.

@@

The rotated loads are not precisely the same as the target astrophysical loads
in part because the target loads only seek to \emph{identify} and \emph{order}
factors that can be interpreted within an astrophysical context. 
\todo{For example,
the first factor does show positive contributions to Al, like the target load,
but it also shows a slight negative contribution with respect to alpha elements
(Mg, Ca)}, and a gradual increase among Fe-peak elements up to Ni. Similarly while
Ca and Mg are indeed positive in the second factor, as targeted, there is a
weak negative contribution to Mn abundances, which was not part of the target
loads. The fourth factor shows a positive contributions to Sr, Y, Ba, and to a
lesser extent, Eu, which is representative of slow neutron capture production.
Similarly, the last component shows a very strong positive contribution towards
Eu, and could be interpreted as production from the rapid nucleosynthesis process.

After accounting for the contributions by all
$J = 5$ factors, the specific scatter remaining in each dimension is shown in
Figure~\ref{fig:exp2-specific-scatter}. The remaining scatter varies from as low 
as 0.08~dex (Mn, Y) up to 0.40~dex (Al, Ba, Ni). Al is not found to have any 
predominant factor in these data: it has non-zero contributions from all loads
(Figure~\ref{fig:exp2-factor-loads}. A similar pattern is seen in Cr, Co, and Ni,
although the specific scatter for Cr and Co is half that of what is found for
Ni (0.2\,dex instead of 0.4\,dex).

% TODO: anything else to say here?

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp2-latent-factors-visualize.pdf}
	\caption{Rotated latent factors found in Experiment~2 using the \citet{Barklem:2005}
			 data. The panels on the left show the entries for each factor load. On
			 the right hand side we show the absolute fractional contributions to
			 each element, which illustrates the loads that contribute most to each
			 element (either by positive or negative factor loads).}
    \label{fig:exp2-factor-loads}
\end{figure*}

%The relatively wide uncertainty regions on these factors is perhaps not
%surprising given that only 61 stars were used in this data set. However, this
%experiment hopefully demonstrates that the rotated latent factors have some
%interpretability.
